Data Specialization for a Staged Lambda Calculus

### 2015-02-27 paper/intro.tex
finished a major pass through the intro (pending the first paragraph)

While performant, this paradigm of programming with explicit collections of
passes results in complex code where invariants must hold across different
passes, and local changes for one pass may require changes to other passes to
``plumb'' inputs from all other frequencies\,\cite{Foley:2011}.
%
In other words, requiring manual separation of passes breaks key programming
abstractions such as composition and modularity.
%
Graphics researchers therefore have suggested using explicitly staged programs
\,\cite{Proudfoot:2001,Foley:2011,He:2014}, deferring to the compiler the
mechanical task of separating such programs into passes. However, prior work in
automatic pass separation has been limited to simple languages that do not
support recursion or first-class functions.

@@

There has been much work on language-based techniques for performing
precomputation. Partial evaluation~\cite{futamura71,jones96} achieves
precomputation by specializing functions to specific arguments. Going back to
our example, if we specialize $f$ to a particular $v$, written $f_v$, such that
$f(v,y) = f_v(y)$, then we can specialize $F$ to $v$ as
\begin{lstlisting}
fun @$F_v(y_0, y_1, \ldots, y_{m-1})$@ = 
  @$f_v(y_0);  f_v(y_1); \ldots; f_v(y_{m-1})$@.
\end{lstlisting}
eliminating the need to recompute
\crem{WIP especially below here}
TODO.
%
A closely related technique, {\em metaprogramming}, allows writing programs that
generate specialized code. Metaprogramming enables fine-grained control over
specialization by requiring staging annotations that make explicit the stage of
each computation. (In partial evaluation, we typically assume that a
\emph{binding-time analysis} has automatically determined which parts of $f$ can
be specialized.)

While previous work shows that precomputation can be performed
essentially automatically (via partial evaluation or meta
programming), frequency reduction has been less explored.  In their
original paper, J{\o}rring and Scherlis proposed {\em pass separation}
as a manual method for {\em splitting} a program into multiple passes.
%
Applied to our example, pass separation would split the function $f$
into $f_1$ and $f_2$ such that $f(x,y) = f_2(f_1(x),y)$.  We can then
rewrite our example as a multi-pass program:
%
\begin{lstlisting}
fun @$F(x, y_0, y_1 \ldots y_{m-1})$@ = 
  let @$fx = f_1(x)$@
  in @$(f_2(fx, y_0);  f_2(fx, y_1)  \ldots f_2(fx, y_{m-1}))$@.
\end{lstlisting}
%
The key difference between partial evaluation (meta programming) and
pass separation is that the latter does not require the first
argument~$x$: it works with all arguments, whereas the partially
evaluated program is specialized for a particular value of $x$ ($x =
[1,2, \ldots, n]$ in the example). Specifically, if desired, partial
evaluation can be applied subsequently in our example, by partially
evaluating function~$f_2$ on $f_1(x)$.

However, unlike for precomputation, there is no available technique
for pass separation except for very simple languages
\cite{knoblock96,Proudfoot:2001,Foley:2011,He:2014}.  Notably omitted
features include recursion and first-class functions.  We are thus
interested in the problem of splitting automatically a functional
language.
%

As our starting point, we consider a modal, typed language \lang\ with explicit
staging in the style of Davies~\cite{davies96}, where the $\fut$
modality denotes computation in a later stage, and present a staged
operational semantics similar for \lang\ (\ref{sec:stagedsemantics}).
We then present a splitting algorithm that pass-separates a \lang\ program
into two monostaged programs, one for the first stage and for the
second stage (\ref{sec:splitting}).  Since each program is monostaged,
they are expressed in conventional functional languages.

The crux of the splitting algorithm is splitting recursive mixed stage
functions that involve both a first stage and second stage
computations.  When split, such a function turns into a first-pass
function that yields a {\em boundary value} that encodes the
``first-stage'' part of the computation.  If the function recurs on a
first-stage argument, then the boundary value is typically a recursive
data structure.  In other words, the splitting algorithm maps
computational recursion into a recursive data type. If the boundary
structure is generated as a result of a conditional, it is typically
encoded as a (recursive) sum type, such as a tree, where the kind of
the node indicates the recursion status of the first-stage evaluation.
The second-pass part of the computation, which corresponds to the
``second stage'' takes the boundary value as an argument as well as
the second-stage argument and uses them to complete the computation,
often by traversing the now precomputed boundary in light of the now
available second-stage argument.

Since it deals with unrestricted recursion and a rich language
consisting of first-class functions, recursion, and sum and recursive
types, the splitting algorithm is naturally complex. With some care,
the algorithm nevertheless can be specified reasonably succinctly. In
fact, we have implemented the algorithm (\ref{sec:implementation}) and
applied it to a number of staged programs (\ref{sec:examples}) ranging
from straight-line arithmetic operations to recursive and higher-order
functions. In the case of \texttt{quickselect}, which we discuss next, 
the split code is asymptotically faster over many $y_i$.

>>>>

While performant, this paradigm of programming with explicit collections of
passes results in complex code where invariants must hold across different
passes, and local changes for one pass may require changes to other passes to
``plumb'' inputs from all other frequencies\,\cite{Foley:2011}.
%
In other words, requiring manual separation of passes breaks key programming
abstractions such as composition and modularity.
%
Graphics researchers therefore have suggested using explicitly staged programs
\,\cite{Proudfoot:2001,Foley:2011,He:2014}, deferring to the compiler the
mechanical task of separating such programs into passes.

@@

There has been much work on language-based techniques for performing
precomputation. Partial evaluation~\cite{futamura71,jones96} achieves
precomputation by specializing functions to specific arguments. Going back to
our example, if we specialize $f$ to a particular $v$, written $f_v$, such that
$f(v,y) = f_v(y)$, then we can specialize $F$ to $v$ as
\begin{lstlisting}
fun @$F_v(y_0, y_1, \ldots, y_{m-1})$@ = 
  @$f_v(y_0);  f_v(y_1); \ldots; f_v(y_{m-1})$@.
\end{lstlisting}
eliminating the need to recompute the parts of $f(v,-)$ which do not depend on
the second argument.

A closely related technique, {\em metaprogramming}, allows writing programs that
generate specialized code. Metaprogramming enables fine-grained control over
specialization by requiring staging annotations that make explicit the stage of
each computation. (In partial evaluation, we typically assume that a
\emph{binding-time analysis} has automatically determined which parts of $f$ can
be specialized.)

While partial evaluation and metaprogramming have been studied extensively,
frequency reduction techniques have not seen the same attention. In
\cite{JS86-staging}, J{\o}rring and Scherlis proposed {\em pass separation} as a
manual method for {\em splitting} a program into multiple passes.
%
In our example, pass separation would split the function $f$ into two passes
$f_1$ and $f_2$ such that $f(x,y_i) = f_2(f_1(x),y_i)$. Then we evaluate $F$ by
evaluating the first pass on $x$, and then the second pass on each $y_i$.
%
\begin{lstlisting}
fun @$F_\textrm{staged}(x, y_0, y_1, \ldots, y_{m-1})$@ = 
  let @$fx = f_1(x)$@
  in @$(f_2(fx, y_0);  f_2(fx, y_1)  \ldots f_2(fx, y_{m-1}))$@.
\end{lstlisting}
%
The key difference between partial evaluation (or metaprogramming) and pass
separation is that the latter can be performed without access to the first
argument $x$: $F_\textrm{staged}$ works for any $x$, while $F_v$ is defined
only for $x=v$. (Of course, when a particular $v$ \emph{is} known, one may
always use partial evaluation to specialize $F_\textrm{staged}$ to that $v$.)

Prior work on partial evaluation and metaprogramming has shown that
precomputation can largely be automated for functional languages. In contrast,
automatic pass separation has been limited to simple languages that do not
support recursion or first-class
functions~\cite{knoblock96,Proudfoot:2001,Foley:2011,He:2014}. In this paper, we
extend this work by presenting a pass separation algorithm for a typed lambda
calculus.

As our starting point, we consider an explicitly-two-staged, typed lambda
calculus \lang\ in the style of Davies~\cite{davies96}, with a $\fut$ modality
denoting computation in a later stage (\ref{sec:semantics}). We then present a
splitting algorithm to pass-separate a \lang\ program into two passes
corresponding to the stages of computation in that program
(\ref{sec:splitting}). These passes are monostaged programs, expressed in a
conventional functional language.

The most interesting case involves splitting a mixed-stage function (like $f$ in
the example) which is moreover recursive. This yields a first pass function
$f_1$ which computes from its first-stage argument $x$ a \emph{boundary} data
structure encoding the interstage communication in $f(x,-)$. When $f$ recurs on
$x$, this boundary structure itself has a recursive type.
%
The second pass is a function $f_2$ which takes the boundary data structure
$f_1(x)$ and a second-stage argument $y_i$, and completes the computation
$f(x,y_i)$, traversing the boundary structure in light of the particular
argument $y_i$.

We have implemented this splitting algorithm (\ref{sec:implementation}) and
applied it to a number of staged programs (\ref{sec:examples}) ranging from
straight-line arithmetic operations to recursive and higher-order functions. In
the case of \texttt{quickselect}, which we discuss next, the split code is in
fact asymptotically faster over many $y_i$.

%The crux of the splitting algorithm is splitting recursive mixed stage
%functions that involve both a first stage and second stage computations.  When
%split, such a function turns into a first-pass function that yields a {\em
%boundary value} that encodes the ``first-stage'' part of the computation.  If
%the function recurs on a first-stage argument, then the boundary value is
%typically a recursive data structure.  In other words, the splitting algorithm
%maps computational recursion into a recursive data type. If the boundary
%structure is generated as a result of a conditional, it is typically encoded as
%a (recursive) sum type, such as a tree, where the kind of the node indicates
%the recursion status of the first-stage evaluation.  The second-pass part of
%the computation, which corresponds to the ``second stage'' takes the boundary
%value as an argument as well as the second-stage argument and uses them to
%complete the computation, often by traversing the now precomputed boundary in
%light of the now available second-stage argument.
